Compositional Certified Resource Bounds\ifshort{\vspace{-.15cm}}{}

### 2014-07-07 paper/main.tex
more typos

\iffull{It might be surprising that we}{We} only track simple linear relations in
the logical context $\Gamma$.  It would of course be possible to keep
of track more sophisticated assertions.  However, this simple form is
sufficient for the examples we considered and we can efficiently
decide queries such as $\Gamma \implies x<y \; ?$ that we make in the
assignment rule.  Nevertheless, this logical part of our analysis
system is orthogonal to the quantitative part and can be easily
extended if necessary.

As mentioned, the automatic analysis can handle challenging example
programs without special tricks or techniques.  Examples
\emph{speed\_1} and \emph{speed\_2} that are taken from previous
work~\cite{GulwaniMC09} and demonstrate the our method can handle
\emph{tricky iteration patterns}.  The SPEED tool~\cite{GulwaniMC09}
derives the same bounds as our analysis but requires inference of
\emph{disjunctive invariants}.  In the abstract interpretation
community, these invariants are known to be notoriously difficult to
handle.
%
In example \emph{speed\_1} we have one loop that first increments
variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
derive the tight bound $|[x, n]| + |[y, m]|$.
%
Example \emph{speed\_2} is more tricky and we found it hard to
manually find a bound.  However, using potential transfer reasoning of
amortized analysis, it is straightforward to prove the tight bound
$|[x, n]| + |[z, n]|$.

Example \emph{t08a} shows the ability of the analysis to discover
interaction between \emph{sequential loops} through size changes of
variables.  We accurately track the size change of $y$ in the first
loop by transferring the potential $0.1$ form $|[y,z]|$ to $|[0,y]|$.
Furthermore, \emph{t08a} shows again that we do not handle the
constants $1$ or $0$ in any special way.  In all examples we could
replace $0$ and $1$ with other constants like in the second loop and
still derive a tight bound.  \iffull{The only information, that the analyzer
needs is $y \geq c$ before assigning $y = y - c$.}{}
%
Example \emph{t27} shows how amortization can be used to handle
\emph{interacting nested loops}.  In the outer loop we increment the
variable $n$ till $n = 0$.  In each of the $|[n,0]|$ iterations, we
increment the variable $y$ by $1000$.  We then non-deterministically
(expressed by \code{*}) execute an inner loop that decrements $y$ by
$100$ until $y<100$.  The analysis discovers that only the first
execution of the inner loop depends on the initial value of $y$.  We
again derive tight constant factors.
%

@@

In this article, we describe our system for a subset of Clight to
discuss general ideas instead of technical details.  This subset is
given by the following grammar.
%
\begin{align*}
S &:= \code{assert}~E
\mid \code{skip}
\mid \code{break}
\mid \code{return}~x
\mid x \gets E
\mid x \gets f(x^*)
\\
& \mid \code{loop}~S
\mid \code{if}(E)~S~\code{else}~S
\mid S;S
\mid \code{tick}(n)
\end{align*}
%
Expressions are $E$ are left abstract in our presentation.  For our
analysis framework, it is only important that they are side-effect
free.
%
The most notable difference to full Clight is that we can only assign
to variables and thus do not consider operations that update the heap.
Moreover, function arguments and return values are assumed to be
variables.  In the implementation we can deal with heap updates and
general function calls and returns.  However, we have not implemented
function pointers, goto statements, continue statements, and switch
statements.

@@

CompCert Clight's operational semantics is based on small-step
transitions and continuations.  Expressions---which do not have side
effects---are evaluated in a big-step fashion.  Here, we describe a
simplified version of Clight's semantics that is sufficient for the
subset that we are describing.

A program state $\state = (\theta, \gamma)$ composed of two maps from
variable names to integers. The first map $\theta : \text{Locals} \to
\Z$ assigns integers to local variables of a function and the second
map $\gamma : \text{Globals} \to \Z$ gives values to global variables
of the program.  In this article, we assume that all values are
integers but in the implementation we support all data types of
Clight.
%
The evaluation function \evalE{\cdot}{} maps an expression $e \in
E$ to a value $\evalE{e}{\state} \in \Z$ in program state $\state$.

For simplicity, we assume that local variables and global variables
are always disjoint.  We just write $\state(x)$ to obtain the value of
$x$ in program state $\state$. \iffull{Such a lookup is defined as
$$
(\theta, \gamma)(x) =
\left\{
\begin{array}{lr}
\theta(x) & \mbox{ if } x \in \dom \theta \\
\gamma(x) & \mbox{ if } x \in \dom \gamma
\end{array}
\right..
$$}{}
\iffull{For a program state $\state$, we}{We} similarly write $\state[x \mapsto v]$
for the program state that maps $x$ to $v$ and behaves as $\state$ for
all other variables, regardless whether $x \in \dom \theta$ or $x \in
\dom \gamma$.

>>>>

\iffull{It might be surprising that we}{We} only track simple linear relations in
the logical context $\Gamma$.  It would of course be possible to keep
track of more sophisticated assertions.  However, this simple form is
sufficient for the examples we considered and we can efficiently
decide queries such as $\Gamma \implies x<y \; ?$ that we make in the
assignment rule.  Nevertheless, this logical part of our analysis
system is orthogonal to the quantitative part and can be easily
extended if necessary.

As mentioned, the automatic analysis can handle challenging example
programs without special tricks or techniques.  Examples
\emph{speed\_1} and \emph{speed\_2} that are taken from previous
work~\cite{GulwaniMC09} and demonstrate that our method can handle
\emph{tricky iteration patterns}.  The SPEED tool~\cite{GulwaniMC09}
derives the same bounds as our analysis but requires heuristics for
its counter instrumentation.  It can also be handled with inference of
\emph{disjunctive invariants}, but in the abstract interpretation
community, these invariants are known to be notoriously difficult to
handle.
%
In example \emph{speed\_1} we have one loop that first increments
variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
derive the tight bound $|[x, n]| + |[y, m]|$.
%
Example \emph{speed\_2} is more tricky and we found it hard to
manually find a bound.  However, using potential transfer reasoning as
in amortized analysis, it is easy to prove the tight bound
$|[x, n]| + |[z, n]|$.

Example \emph{t08a} shows the ability of the analysis to discover
interaction between \emph{sequenced loops} through size change of
variables.  We accurately track the size change of $y$ in the first
loop by transferring the potential $0.1$ form $|[y,z]|$ to $|[0,y]|$.
Furthermore, \emph{t08a} shows again that we do not handle the
constants $1$ or $0$ in any special way.  In all examples we could
replace $0$ and $1$ with other constants like in the second loop and
still derive a tight bound.  \iffull{The only information, that the analyzer
needs is $y \geq c$ before assigning $y = y - c$.}{}
%
Example \emph{t27} shows how amortization can be used to handle
\emph{interacting nested loops}.  In the outer loop we increment the
variable $n$ until $n = 0$.  In each of the $|[n,0]|$ iterations, we
increment the variable $y$ by $1000$.  Then we non-deterministically
(expressed by \code{*}) execute an inner loop that decrements $y$ by
$100$ until $y<100$.  The analysis discovers that only the first
execution of the inner loop depends on the initial value of $y$.  We
again derive tight constant factors.
%

@@

In this article, we describe our system for a subset of Clight to
discuss general ideas instead of technical details.  This subset is
given by the following grammar.
%
\begin{align*}
S &:= \code{assert}~E
\mid \code{skip}
\mid \code{break}
\mid \code{return}~x
\mid x \gets E
\mid x \gets f(x^*)
\\
& \mid \code{loop}~S
\mid \code{if}(E)~S~\code{else}~S
\mid S;S
\mid \code{tick}(n)
\end{align*}
%
Expressions $E$ are left abstract in our presentation.  For our
analysis framework, it is only important that they are side-effect
free.
%
The most notable difference to full Clight is that we can only assign
to variables and thus do not consider operations that update the heap.
Moreover, function arguments and return values are assumed to be
variables.  In the implementation we can deal with heap updates and
general function calls and returns.  However, we have not implemented
function pointers, \code{goto} statements, \code{continue} statements,
and \code{switch} statements.

@@

CompCert Clight's operational semantics is based on small-step
transitions and continuations.  Expressions---which do not have side
effects---are evaluated in a big-step fashion.  Here, we describe a
simplified version of Clight's semantics for the subset we are
describing.

A program state $\state = (\theta, \gamma)$ composed of two maps from
variable names to integers. The first map $\theta : \text{Locals} \to
\Z$ assigns integers to local variables of a function and the second
map $\gamma : \text{Globals} \to \Z$ gives values to global variables
of the program.  In this article, we assume that all values are
integers but in the implementation we support all data types of
Clight.
%
The evaluation function \evalE{\cdot}{} maps an expression $e \in
E$ to a value $\evalE{e}{\state} \in \Z$ in the program state
$\state$.

For simplicity, we assume that local variables and global variables
are always different.  We just write $\state(x)$ to obtain the value of
$x$ in program state $\state$. \iffull{Such a lookup is defined as
$$
(\theta, \gamma)(x) =
\left\{
\begin{array}{lr}
\theta(x) & \mbox{ if } x \in \dom \theta \\
\gamma(x) & \mbox{ if } x \in \dom \gamma
\end{array}
\right..
$$}{}
\iffull{For a program state $\state$, we}{We} similarly write $\state[x \mapsto v]$
for the program state that maps $x$ to $v$ and behaves as $\state$ for
all other variables, regardless whether $x \in \dom \theta$ or $x \in
\dom \gamma$.
